---
title: "INFO6105_FinalProject"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# **📘 Introduction**

### **Context & Motivation**

-   GitHub is one of the most widely used platforms for software
    development, open-source collaboration, and community-driven
    innovation.

-   “Engagement” on GitHub — measured through watchers, stars, forks,
    and activity — serves as a key signal of visibility, user interest,
    and long-term project sustainability.

-   Understanding **what drives engagement** is useful for developers,
    maintainers, and organizations aiming to build successful
    repositories.

### **Analytical Goal**

-   The central goal of this project is to quantify how repository
    characteristics relate to engagement levels across GitHub projects.

-   Specifically, the project aims to identify:

    -   Which **numerical features** (stars, forks, age) most strongly
        predict engagement.

    -   Whether **programming language** contributes to systematic
        differences in engagement.

    -   Whether **language and popularity interact** — i.e., does
        language matter *more* for popular repositories?

    ### **Summary of Approach**

    -   A Kaggle dataset of **215,029 GitHub repositories** was cleaned
        and processed.

    -   A random sample of **500 repositories** was used to perform
        formal statistical modeling.

    -   Three analyses were carried out:

        1.  **Multiple Linear Regression** — to quantify predictors of
            engagement.

        2.  **One-Way ANOVA** — to test differences in engagement by
            language.

        3.  **Two-Way ANOVA** — to test language, popularity, and
            interaction effects.

    -   Results contribute to a clearer understanding of what drives
        community interest on GitHub.

# **Step 1: Import Libraries**

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(knitr)
library(broom)
library(car)
library(ggpubr)

data <- read_csv("Data/rgithub_repos.csv")
data <- clean_names(data)
glimpse(data)


```

# **Step 2: Data Cleaning + Feature Engineering**

```{r}
## ---- data_cleaning, message=FALSE, warning=FALSE ----

# Remove duplicated rows if they exist
data <- data %>% distinct()

# Convert character columns to factors where appropriate
data$language <- as.factor(data$language)
data$license <- as.factor(data$license)
data$default_branch <- as.factor(data$default_branch)

# Convert logical columns to numeric (TRUE=1, FALSE=0)
logical_cols <- sapply(data, is.logical)
data[logical_cols] <- lapply(data[logical_cols], as.numeric)

# Create repo age in years
data$created_at <- as.POSIXct(data$created_at, format="%Y-%m-%dT%H:%M:%S")
data$updated_at <- as.POSIXct(data$updated_at, format="%Y-%m-%dT%H:%M:%S")
data$repo_age_years <- as.numeric(difftime(Sys.time(), data$created_at, units = "days"))/365

# Replace missing values in numeric columns with median
num_cols <- sapply(data, is.numeric)
data[num_cols] <- lapply(data[num_cols], function(x){
  x[is.na(x)] <- median(x, na.rm=TRUE)
  x
})

# Replace missing language values with "Unknown"
data$language <- fct_explicit_na(data$language, na_level = "Unknown")

# Create language groups for ANOVA
data$language_group <- ifelse(data$language %in% c("Python", "JavaScript", "TypeScript"),
                              as.character(data$language),
                              "Other") %>% 
  factor()

# Create popularity groups for t-test & regression
median_stars <- median(data$stars, na.rm=TRUE)
data$popularity_group <- ifelse(data$stars > median_stars, "High Stars", "Low Stars") %>% 
  factor()

# Sample 500 rows for analysis
set.seed(6105)
sample_data <- data %>% sample_n(500)

glimpse(sample_data)

```

```{r}
names(sample_data)
```

```{r eda_summary, message=FALSE, warning=FALSE}
library(psych)   # for describe()

# Summary statistics (only existing numeric columns)
describe(sample_data[, c("stars",
                         "forks",
                         "watchers",
                         "issues",
                         "size",
                         "repo_age_years")])

```

# **Step 3: Exploratory Data Analysis**

```{r eda_plots, message=FALSE, warning=FALSE}
# 1. Distribution of Stars
ggplot(sample_data, aes(stars)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Repository Stars",
       x = "Stars", y = "Count")

# 2. Stars vs Forks (Scatter)
ggplot(sample_data, aes(stars, forks)) +
  geom_point(alpha = 0.6, color = "darkred") +
  labs(title = "Stars vs Forks",
       x = "Stars", y = "Forks")

# 3. Watchers by Popularity Group
ggplot(sample_data, aes(popularity_group, watchers, fill = popularity_group)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Watchers by Popularity Group",
       x = "Popularity Group", y = "Watchers") +
  theme(legend.position = "none")

# 4. Stars by Language Group
ggplot(sample_data, aes(language_group, stars, fill = language_group)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Stars by Language Group",
       x = "Language Group", y = "Stars") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r transform_vars, message=FALSE}
# Log-transform highly skewed variables
sample_data <- sample_data %>%
  mutate(
    log_stars     = log10(stars + 1),
    log_forks     = log10(forks + 1),
    log_watchers  = log10(watchers + 1)
  )

# Quick check
summary(sample_data[, c("log_stars", "log_forks", "log_watchers", "repo_age_years")])

```

## 2. Analysis 1 – Multiple Linear Regression

Response: log_watchers (proxy for engagement)\
Predictors: log_stars, log_forks, repo_age_years

```{r mlr_fit, message=FALSE}
# Keep only complete cases for the regression
mlr_data <- sample_data %>%
  select(log_watchers, log_stars, log_forks, repo_age_years) %>%
  drop_na()

# Fit multiple linear regression
model_mlr <- lm(log_watchers ~ log_stars + log_forks + repo_age_years,
                data = mlr_data)

# Coefficients, p-values, etc.
broom::tidy(model_mlr)

# Overall fit (R², adj R², F-test)
broom::glance(model_mlr)
```

```{r}
## ---- mlr_diagnostics, message=FALSE, warning=FALSE ----
library(ggplot2)

# Extract model data
mlr_data$predicted <- predict(model_mlr)
mlr_data$residuals <- resid(model_mlr)

# 1. Residuals vs Fitted (checks linearity & equal variance)
ggplot(mlr_data, aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals")

# 2. Q-Q Plot (normality of residuals)
ggplot(mlr_data, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Residuals")

# 3. Scale-Location Plot (homoscedasticity)
ggplot(mlr_data, aes(x = predicted, y = sqrt(abs(residuals)))) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(se = FALSE) +
  labs(title = "Scale-Location Plot",
       x = "Fitted Values",
       y = "√|Residuals|")
```

## 3. Analysis 2 – One-Way ANOVA

Research Question:\
Do average engagement levels (log_watchers) differ across primary
language groups for popular GitHub repositories?

Response: log_watchers (proxy for engagement)\
Factor: language_group (JavaScript, Python, TypeScript, Other)

Null hypothesis (H₀): Mean log_watchers is the same for all language
groups.\
Alternative hypothesis (H₁): At least one language group has a different
mean log_watchers.

```{r anova1_descriptives, message=FALSE, warning=FALSE}
# Keep only rows with both variables present
anova1_data <- sample_data %>%
  drop_na(language_group, log_watchers)

# Summary statistics by language group
anova1_summary <- anova1_data %>%
  group_by(language_group) %>%
  summarise(
    n = n(),
    mean_log_watchers = mean(log_watchers),
    sd_log_watchers   = sd(log_watchers)
  )

knitr::kable(anova1_summary, digits = 3,
             caption = "Summary statistics of log_watchers by language group")

```

🔹 3.1 **Descriptive Stats & Boxplot**

```{r anova1_boxplot, message=FALSE, warning=FALSE}
# Boxplot of engagement by language
ggplot(anova1_data,
       aes(x = language_group, y = log_watchers, fill = language_group)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Engagement (log_watchers) by Language Group",
       x = "Language Group",
       y = "log(Watchers)") +
  theme(legend.position = "none")
```

🔹 3.2 **One-Way ANOVA**

```{r anova1_model, message=FALSE, warning=FALSE}
# Fit one-way ANOVA: engagement ~ language_group
fit_aov1 <- aov(log_watchers ~ language_group, data = anova1_data)

# ANOVA table
broom::tidy(fit_aov1)
```

## **Interpretation of One-Way ANOVA**

The one-way ANOVA tested whether average engagement levels (measured
using *log-transformed watchers*) differ across programming language
groups. Results indicate a **statistically significant effect of
language group on engagement**, **F(3, 496) = 4.76, p = 0.0028**. This
means that *at least one* language group has a different average
log_watchers value compared to the others. In simpler terms, **the
popularity of repositories differs depending on the primary programming
language.**

🔹 3.3 **Tukey HSD Post-hoc Test**s

```{r anova1_tukey, message=FALSE, warning=FALSE}
tukey_aov1 <- TukeyHSD(fit_aov1)

# Full Tukey output
tukey_aov1

# Tidy version (nice for tables)
tukey_tidy <- broom::tidy(tukey_aov1)
knitr::kable(tukey_tidy, digits = 3,
             caption = "Tukey HSD pairwise comparisons for language_group")
```

## **Tukey HSD Interpretation**

Because the ANOVA was significant, we performed Tukey HSD post-hoc tests
to identify which pairs differ.

Key findings:

-   **Python vs JavaScript** showed one of the strongest significant
    differences

    (p ≈ 0.0014), indicating that Python repositories have **lower**
    engagement than JavaScript repositories on average.

-   **Python vs Other** was also statistically significant (p ≈ 0.0379),
    suggesting Python tends to have lower engagement than the “Other”
    category.

-   **All other pairwise comparisons** (e.g., TypeScript vs JavaScript,
    TypeScript vs Python, etc.)

    were **not statistically significant**, meaning their mean
    engagement levels are not reliably different.

Overall, **JavaScript repositories generally exhibit higher
engagement**, while **Python tends to show lower engagement** within
this sample.

🔹 3.4 Effect Size (η²)

```{r anova1_effect_size, message=FALSE, warning=FALSE}
anova_table <- broom::tidy(fit_aov1)

# Total SS = sum of all sums of squares
ss_total <- sum(anova_table$sumsq)

# Eta-squared for language_group
eta2_language <- anova_table$sumsq[anova_table$term == "language_group"] / ss_total
eta2_language
```

## **Effect Size Interpretation (η²)**

The effect size (eta-squared) was:

**η² = 0.0279**

This indicates that **only about 2.8% of the variance** in engagement
(log_watchers) is explained by the programming language used.

According to conventional effect-size guidelines:

-   0.01 = small

-   0.06 = medium

-   0.14 = large

This effect size is **small**, meaning that although language group
matters statistically, it is **not a strong predictor** of engagement.

Most variation in engagement is driven by other factors (e.g., stars,
forks, repo age).

🔹 3.5 Assumption Checks (Normality & Equal Variance)

```{r anova1_diagnostics, message=FALSE, warning=FALSE}
# Data frame of fitted values and residuals
aov1_diag <- data.frame(
  fitted     = fitted(fit_aov1),
  residuals  = resid(fit_aov1)
)

# 1. Residuals vs Fitted (check equal variances & linearity of group means)
ggplot(aov1_diag, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "ANOVA 1: Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals")

# 2. Q-Q Plot (normality of residuals)
ggplot(aov1_diag, aes(sample = residuals)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line(color = "red") +
  labs(title = "ANOVA 1: Q-Q Plot of Residuals")

# 3. Scale-Location (homoscedasticity)
ggplot(aov1_diag, aes(x = fitted, y = sqrt(abs(residuals)))) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  labs(title = "ANOVA 1: Scale-Location Plot",
       x = "Fitted Values",
       y = "√|Residuals|")

# 4. Levene's test for equal variances
car::leveneTest(log_watchers ~ language_group, data = anova1_data)
```

It appears that residuals are extremely close to zero due to perfect
correlation among predictors, therefore diagnostic plots appear flat.
The ANOVA residual plots look normal and Levene’s test shows significant
differences.

## **Assumptions Check Interpretation**

### **Normality of Residuals**

The Q-Q plot shows points that generally follow the diagonal line,
though with some deviations at the extremes.

Given the large sample size (n = 500), these deviations are not serious.

Normality is **reasonably satisfied**.

### **Homogeneity of Variance**

Levene’s Test returned:

**F = 5.72, p = 0.00074**

Since **p \< 0.05**, the variances across language groups are **not
equal**.

This violates the homogeneity assumption of ANOVA.

However, ANOVA is fairly robust to variance inequality when group sizes
are similar and sample size is large — which is true for this dataset.

Thus, results remain interpretable but should be described as
**cautiously interpreted**.

### **Residuals vs Fitted & Scale-Location**

-   Residuals appear very close to zero because the transformed
    variables are highly correlated.

-   The Scale-Location plot does not show strong patterns of
    heteroscedasticity.

-   No major violations of linearity appear.

Overall, **assumptions are reasonably satisfied**, with the exception of
unequal variances.

# **4. Analysis 3: Two-Way ANOVA**

**Research Question:**

Do engagement levels differ by **programming language**, by
**popularity**, and is there an **interaction** between them?

### **▣ Factors**

-   **Factor A:** language_group (4 levels)

-   **Factor B:** popularity_group (High Stars / Low Stars)

-   **Interaction:** A × B

-   **Response:** log_watchers

    # **🔹4.1 Create Dataset**

    ```{r}
    anova2_data <- sample_data %>%
      drop_na(language_group, popularity_group, log_watchers)
    ```

# **🔹 4.2 Interaction Plot**

```{r}
interaction.plot(
  x.factor = anova2_data$language_group,
  trace.factor = anova2_data$popularity_group,
  response = anova2_data$log_watchers,
  fun = mean,
  type = "b",
  col = c("red", "blue"),
  pch = c(19, 17),
  xlab = "Language Group",
  ylab = "Mean log(Watchers)",
  trace.label = "Popularity Group",
  main = "Interaction Plot: Language × Popularity"
)
```

# **🔹4.3 Two-Way ANOVA Model**

```{r}
fit_aov2 <- aov(
  log_watchers ~ language_group * popularity_group,
  data = anova2_data
)

anova2_results <- broom::tidy(fit_aov2)
knitr::kable(anova2_results, digits = 3,
             caption = "Two-Way ANOVA Table")
```

# **🔹4.4 Interpretation**

The two-way ANOVA revealed **three clear findings**:

### **1. Main Effect of Programming Language — Significant**

Language group had a **statistically significant effect** on engagement
(p \< .001).

Some languages consistently attract more watchers than others, even
after controlling for popularity.

In particular, JavaScript and TypeScript tend to draw higher engagement
than Python.

### **2. Main Effect of Popularity — Highly Significant**

Popularity group (High Stars vs Low Stars) showed an **extremely strong
effect** (p \< .001).

High-star repositories receive substantially higher engagement
regardless of language.

This confirms that **popularity is the dominant driver of watcher
counts**.

### **3. Interaction Effect — Significant**

The interaction between language and popularity was **significant** (p =
.012), meaning:

> **The influence of programming language on engagement depends on
> whether the repository is popular.**

From the interaction plot:

-   Among **high-star** repositories, language differences are
    noticeable

    (JavaScript \> Other \> Python).

-   Among **low-star** repositories, engagement is uniformly low across
    all languages.

Thus, language matters **only when repositories are already popular**.

# **🔹4.5 Effect Size (partial η²)**

```{r}
## ---- anova2_effect_size, message=FALSE, warning=FALSE ----
anova2_raw <- summary(fit_aov2)

# Extract SS values
ss_language   <- anova2_raw[[1]]["language_group", "Sum Sq"]
ss_popularity <- anova2_raw[[1]]["popularity_group", "Sum Sq"]
ss_interaction <- anova2_raw[[1]]["language_group:popularity_group", "Sum Sq"]
ss_residuals  <- anova2_raw[[1]]["Residuals", "Sum Sq"]

# Partial eta-squared
eta2_language_partial   <- ss_language / (ss_language + ss_residuals)
eta2_popularity_partial <- ss_popularity / (ss_popularity + ss_residuals)
eta2_interaction_partial <- ss_interaction / (ss_interaction + ss_residuals)

eta2_language_partial
eta2_popularity_partial
eta2_interaction_partial
```

Popularity has a **dominant influence** on engagement, explaining **over
53%** of the variance in log-watchers. far more than any other factor.

The interaction between language and popularity explains **\~2%** of the
variance, indicating a **weak but noticeable interaction pattern**.

The effect of language alone is minimal, aligning with earlier findings
that language is not a major determinant of engagement once popularity
is accounted for.

# **🔹4.6 Post-Hoc Tests**

```{r}
## ---- anova2_posthoc, message=FALSE, warning=FALSE ----

# Split data by popularity group
high_data <- subset(anova2_data, popularity_group == "High Stars")
low_data  <- subset(anova2_data, popularity_group == "Low Stars")

# Tukey per popularity level
tukey_high <- TukeyHSD(aov(log_watchers ~ language_group, data = high_data))
tukey_low  <- TukeyHSD(aov(log_watchers ~ language_group, data = low_data))

tukey_high
tukey_low
```

Post-hoc tests show that **language differences only appear in the top
half of repositories** (High Stars group), and even there, the effect is
small.

Among low-engagement projects, **all languages perform similarly**,
confirming that language matters little when a project is not already
popular.

# **🔹4.7 Assumption Checks**

### **Residuals + Fitted**

```{r}
aov2_diag <- data.frame(
  fitted = fitted(fit_aov2),
  residuals = resid(fit_aov2)
)

ggplot(aov2_diag, aes(fitted, residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color="red") +
  labs(title="Two-Way ANOVA: Residuals vs Fitted")
```

No major violations of linearity. Slight heteroscedasticity but
acceptable with n = 500.

### **Normal Q-Q**

```{r}
ggplot(aov2_diag, aes(sample=residuals)) +
  stat_qq() +
  stat_qq_line(color="red") +
  labs(title="Two-Way ANOVA: Q-Q Plot")
```

Residuals are approximately normal. Minor tail deviations do not
threaten ANOVA validity with large samples.

### **Scale-Location**

```{r}
ggplot(aov2_diag, aes(fitted, sqrt(abs(residuals)))) +
  geom_point(alpha=0.6) +
  geom_smooth(se=FALSE) +
  labs(title="Two-Way ANOVA: Scale-Location Plot")
```

Mild heteroscedasticity is present, but ANOVA is robust given balanced
groups and sample size.

### **Levene’s Test**

```{r}
leveneTest(log_watchers ~ language_group * popularity_group, 
           data = anova2_data)
```

Assumption of equal variance is violated.

However, with:

-   large sample (n = 500),

-   balanced groups,

-   robust ANOVA models,

the results remain valid but should be described as **interpreted with
caution**.

# **Summary & Insights**

This project examined engagement patterns across GitHub repositories
using a Kaggle dataset of over 215,000 entries, with a random sample of
500 used for statistical analyses. Engagement was measured using
log-transformed watcher counts, and the study explored how repository
features—particularly stars, forks, age, and programming language—shape
engagement.

### **Multiple Linear Regression Findings**

The regression analysis revealed that **stars are by far the strongest
predictor of engagement**, showing a near-perfect linear relationship
with log-watchers. Forks and repository age added minimal explanatory
power once stars were included, highlighting significant
multicollinearity among GitHub popularity indicators (stars, forks,
watchers). The model’s extremely high R² value reflects this overlap,
reinforcing that engagement on GitHub is overwhelmingly tied to existing
popularity signals rather than to structural attributes like age.

### **One-Way ANOVA Results (Language Effects)**

A one-way ANOVA tested whether engagement varied across programming
language groups. Results showed a **statistically significant but modest
difference** in engagement by language (F(3, 496) = 4.76, p = 0.0028).

Post-hoc Tukey comparisons indicated:

-   **JavaScript \> Python**

-   **Other \> Python**

However, the effect size was small (η² = 0.028), meaning language
explained only about **3% of the variation** in engagement. Thus, while
statistically detectable, the practical importance of language is
limited.

### **Assumption Checks**

-   **Normality:** Residuals were approximately normal.

-   **Equal Variance:** Levene’s test indicated unequal variances (p \<
    0.001), but with balanced groups and a large sample size, ANOVA
    results remain reliable.

-   **Linearity & Homoscedasticity:** Plots showed acceptable patterns
    with no major violations.

### **Overall Insights**

Although programming language shows minor statistical differences, **its
practical impact on engagement is minimal**. The dominant driver of
engagement is **popularity itself**, primarily captured through stars.
Repositories that have already accumulated attention regardless of
language tend to attract more watchers, while forks and age play
comparatively small roles.

## **Final Conclusion**

Overall, the analysis demonstrates that **GitHub engagement is driven
far more by existing popularity than by technical characteristics of a
project**. Stars provide a strong and direct signal of community
interest, making them the clearest predictor of engagement.

Programming language contributes only modestly to engagement patterns,
suggesting that developers should focus less on language choice and more
on building visibility, maintaining activity, and attracting early
adoption.

```{r}
sessionInfo()
```
